Supplementary Information

### 2021-06-03 doc/manuscript/ForestGEO_treerings.Rmd
revamp abstract + minor cleanup
#121

Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape individual tree growth and forest productivity.
However, conventional tree-ring analysis methods are not designed to simultaneously account for the effects of climate, tree size, and other environmental drivers on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
Interactions between DBH and climate were common (44% of cases tested). 
Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
Accounting for DBH, growth rate declined over time for 90% of species-site combinations.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing tree growth than has previously been possible. 
*As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent, we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.*

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major unresolved questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.

\newpage
**Table 1 | Summary of hypotheses and specific predictions tested here, along with the frenquency at which they were supported**.  
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypotheses_table <- hypotheses_table[which(hypotheses_table$include==1),c(2:7)]
question_rows<- which(!is.na(hypotheses_table$Q) )
hypothesis_rows<-which(!is.na(hypotheses_table$H) )

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
The most influential seasonal windows were most commonly long ($\ge$3 months at 7 of the 9 sites with significant main effects) and coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 7 sites with significant main effects).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
Outside of the tropics, the most influential windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

Within the temperature group (Fig. `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$3 months at 9 of 10 sites).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

>>>>

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously account for the effects of climate, tree size, and other drivers on individual growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually over $\le$ 3-month seasonal windows), with both included in 78% of top models, and with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested). 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently peaked at intermediate DBH and declined thereafter.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing tree growth than has previously been possible.

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.

\newpage
**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported**.  
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypotheses_table <- hypotheses_table[which(hypotheses_table$include==1),c(2:7)]
question_rows<- which(!is.na(hypotheses_table$Q) )
hypothesis_rows<-which(!is.na(hypotheses_table$H) )

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
The most influential seasonal windows were most commonly long ($\ge$ 3 months at 7 of the 9 sites with significant main effects) and coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 7 sites with significant main effects).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
Outside of the tropics, the most influential windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

Within the temperature group (Fig. `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$ 3 months at 9 of 10 sites).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).
